Reduced plasticity and opportunity for selection in terrestrial ectotherm populations under climate change

### 2023-12-11 ms.qmd
more writing

Climate change is expected to result in warmer and more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability in the past is expected to have resulted in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016]. The extent to which adaptive plasticity and genetic adaptation will allow organisms to adapt to new environmental conditions is not yet clearly established [@Chevin2017]. Without such responses, high rates of extinction are predicted if organisms cannot track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to be the 'first line of defence' against changing climates by buying time for genetic adaptation to take place [i.e., the 'plasticty first hypothesis' - @WestEberhard2003; see also @Lande2009]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity is low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998]. Despite this theoretical expectation, empirical support is scant [], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015; @Dewitt1998], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity) or life-history stages.

@@

Here, we use meta-analysis and new effect sizes to re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity. We then expand these effect sizes to allow for comparisons of variance in physiological rates to ask the following questions: 1) How much is variance in physiological rates expected to change, if at all, as temperatures rise? 2) Are temperture effects on plastic adjustments in physiolgical rates larger than changes in variance across aquatic and terrestrial ectotherms? 3) Are changes in plasticity or variance in physiological rates impacted by a population's past climate history? and 4) How are means and variances in physiological rates expected to change under climate change?

@@

### *Past climate variability does not influence acclimation capacity or expected change in variance*

@@

Predictions of current global changes in the opportunity for selection were generally conservative with our model explaining ~ 50% of the variation in the observed data ($R^2$ = `r round(mean(r2_2), 2)`, 95% CI: `r round(quantile(r2_2, 0.025)[1], 2)` to `r round(quantile(r2_2, 0.975)[1], 2)`). Across habitat types climate change is predicted to result in a only a `r lapply(mean_perc_habitat, mean)$f`% change in variance for freshwater systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$f)`), whereas we expect a `r abs(lapply(mean_perc_habitat, mean)$m)`% reduction in variance for marine systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$m)`) and a `r abs(lapply(mean_perc_habitat, mean)$t)`% reduction in variance for terrestrial systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[2]`%, $p_{MCMC}$ = `r p_value(pmcmc(mean_perc_habitat$t))`) under a RCP8.5 climate scenerio.

@@

>>>>

Climate change is expected to result in warmer and more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability in the past is expected to have resulted in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016]. The extent to which adaptive plasticity and genetic adaptation will allow organisms to adapt to new environmental conditions is not yet clearly established [@Chevin2017]. Without such responses, high rates of extinction are predicted if organisms cannot track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to be the 'first line of defence' against changing climates by buying time for genetic adaptation to take place [i.e., the 'plasticty first hypothesis' - @WestEberhard2003; see also @Lande2009]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity is low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Despite this theoretical expectation, empirical support is scant [but see @Leung2020], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015; @Dewitt1998], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity) or life-history stages.

@@

Here, we use new effect sizes and meta-analysis to re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity. We then expand these effect sizes to allow for comparisons of variance in physiological rates to ask the following questions: 1) How much is variance in physiological rates expected to change, if at all, as temperatures rise? 2) Are temperture effects on plastic adjustments in physiolgical rates larger than changes in variance across aquatic and terrestrial ectotherms? 3) Are changes in plasticity or variance in physiological rates impacted by a population's past climate history? and 4) How are means and variances in physiological rates expected to change under climate change?

@@

### *Past climate does not influence acclimation capacity or expected change in variance*

@@

Predictions of current global changes in the opportunity for selection were generally conservative with our model explaining ~ 50% of the variation in the observed data ($R^2$ = `r round(mean(r2_2), 2)`, 95% CI: `r round(quantile(r2_2, 0.025)[1], 2)` to `r round(quantile(r2_2, 0.975)[1], 2)`). Across habitat types climate change is predicted to result in a only a `r abs(lapply(mean_perc_habitat, mean)$f)`% change in variance for freshwater systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$f)`), whereas we expect a `r abs(lapply(mean_perc_habitat, mean)$m)`% reduction in variance for marine systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$m)`) and a `r abs(lapply(mean_perc_habitat, mean)$t)`% reduction in variance for terrestrial systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[2]`%, $p_{MCMC}$ = `r p_value(pmcmc(mean_perc_habitat$t))`) under a RCP8.5 climate scenerio.

@@

### **Acclimation capacity is limited in ectotherms**
